### 2021-04-09 ch01_introduction/introduction.tex
Used lowercase versions of namecref(s) at correct places

This thesis aims to reduce uncertainties in multi-model climate projections
with observations by addressing the following three
\namecrefs{enum:01:question_1}:

@@

Parts of this thesis are already published in multiple peer-reviewed
publications (two first-author studies and six co-author studies). A complete
list of these publications is given on \cpageref{ch:bib:own}. Wherever material
from these studies is presented in this thesis, the pronoun \enquote{we} is
used to increase readability by avoiding the passive voice and to acknowledge
all involved contributors. However, unless stated otherwise, all content from
these publications (text, figures, and tables) shown in this thesis originates
from the author of this thesis. A detailed list of contributions to these
studies is given in the corresponding \namecrefs{ch:03:esmvaltool}.

This thesis is structured as follows: \Cref{ch:02:scientific_background}
introduces the scientific background. This includes relevant literature that is
used as a baseline for this thesis. \Cref{ch:03:esmvaltool} gives an overview
of the contributions made to the \ac{ESMValTool}, an open-source software
package for the analysis of \acp{ESM}. These contributions helped to improve
the routine evaluation of \acp{ESM} which is useful for the entire scientific
community and led to co-authorship in four peer-reviewed studies
\autocite{Eyring2020, Lauer2020, Righi2020, Weigel2020}.
\Cref{ch:04:papers_ecs_tcr_assessment} covers the assessment of climate
sensitivity metrics like the \ac{ECS} or \ac{TCR} in the latest generation of
\acp{ESM} from \acs{CMIP}6. This work is already published in two scientific
publications \autocite{Bock2020, Meehl2020}. Since the \ac{ECS} and \ac{TCR}
are considerably higher in this new climate model generation,
\cref{ch:05:paper_ecs} describes the assessment of emergent constraints on the
\ac{ECS} for these \acs{CMIP}6 models and compares these to results derived
from \acs{CMIP}5 models. The content of this \namecref{ch:05:paper_ecs} is
published in \emph{Earth System Dynamics} \autocite{Schlund2020a}.
\Cref{ch:06:paper_gpp} focuses on a new method to reduce uncertainties in
multi-dimensional (gridded) multi-model projections of the future climate with
observations based on \ac{ML}. As an example, the method is applied to \ac{GPP}
at the end of the \nth{21} century, which is already published in the
\emph{Journal of Geophysical Research: Biogeosciences} \autocite{Schlund2020}.
Finally, \cref{ch:07:conclusion} provides a summary of the results of this
thesis and gives an outlook of possible future work.

>>>>

This thesis aims to reduce uncertainties in multi-model climate projections
with observations by addressing the following three
\lcnamecrefs{enum:01:question_1}:

@@

Parts of this thesis are already published in multiple peer-reviewed
publications (two first-author studies and six co-author studies). A complete
list of these publications is given on \cpageref{ch:bib:own}. Wherever material
from these studies is presented in this thesis, the pronoun \enquote{we} is
used to increase readability by avoiding the passive voice and to acknowledge
all involved contributors. However, unless stated otherwise, all content from
these publications (text, figures, and tables) shown in this thesis originates
from the author of this thesis. A detailed list of contributions to these
studies is given in the corresponding \lcnamecrefs{ch:03:esmvaltool}.

This thesis is structured as follows: \Cref{ch:02:scientific_background}
introduces the scientific background. This includes relevant literature that is
used as a baseline for this thesis. \Cref{ch:03:esmvaltool} gives an overview
of the contributions made to the \ac{ESMValTool}, an open-source software
package for the analysis of \acp{ESM}. These contributions helped to improve
the routine evaluation of \acp{ESM} which is useful for the entire scientific
community and led to co-authorship in four peer-reviewed studies
\autocite{Eyring2020, Lauer2020, Righi2020, Weigel2020}.
\Cref{ch:04:papers_ecs_tcr_assessment} covers the assessment of climate
sensitivity metrics like the \ac{ECS} or \ac{TCR} in the latest generation of
\acp{ESM} from \acs{CMIP}6. This work is already published in two scientific
publications \autocite{Bock2020, Meehl2020}. Since the \ac{ECS} and \ac{TCR}
are considerably higher in this new climate model generation,
\cref{ch:05:paper_ecs} describes the assessment of emergent constraints on the
\ac{ECS} for these \acs{CMIP}6 models and compares these to results derived
from \acs{CMIP}5 models. The content of this \lcnamecref{ch:05:paper_ecs} is
published in \emph{Earth System Dynamics} \autocite{Schlund2020a}.
\Cref{ch:06:paper_gpp} focuses on a new method to reduce uncertainties in
multi-dimensional (gridded) multi-model projections of the future climate with
observations based on \ac{ML}. As an example, the method is applied to \ac{GPP}
at the end of the \nth{21} century, which is already published in the
\emph{Journal of Geophysical Research: Biogeosciences} \autocite{Schlund2020}.
Finally, \cref{ch:07:conclusion} provides a summary of the results of this
thesis and gives an outlook of possible future work.

### 2021-04-09 ch02_scientific_background/scientific_background.tex
Used lowercase versions of namecref(s) at correct places

This \namecref{ch:02:scientific_background} introduces the scientific
background of this thesis. First, basic concepts of climate model simulations
and associated uncertainties are introduced (\cref{sec:02:esms}). Next,
important metrics describing climate sensitivity
(\cref{sec:02:climate_sensitivity}) and fundamental biogeochemical processes of
the global carbon cycle (\cref{sec:02:esms}) are presented. Finally,
state-of-the-art techniques that can be used to reduce uncertainties in
projections of the future climate with observations are shown
(\cref{sec:02:reducing_uncertainties}). These methods form the basis for new
techniques developed in this thesis.

@@

In this thesis, climate model data from the two most recent \ac{CMIP}
generations is used, \acs{CMIP}5 and \acs{CMIP}6. More detailed information
about the specific variables and experiments analyzed is given in the
corresponding \namecrefs{ch:04:papers_ecs_tcr_assessment}.

@@

Currently, only about $50 \unit{\%}$ of the emitted \ac{CO2} by humankind
remains in the atmosphere \autocite{Friedlingstein2020}. The residual part is
absorbed by the two other main carbon sinks of the planet: the terrestrial
biosphere and the ocean. These exchange fluxes of carbon are part of the global
carbon cycle of planet Earth. Since only atmospheric \ac{CO2} can act as
\ac{GHG}, this removal of \ac{CO2} from the atmosphere substantially slows down
global warming. Moreover, since the emission of carbon-based \acp{GHG}
introduces additional carbon into the Earth system, the carbon exchange fluxes
might change under global warming, which in turn directly influences the
magnitude of climate sensitivity. However, the two idealized climate
sensitivity metrics \ac{ECS} and \ac{TCR} introduced in the previous
\namecref{sec:02:climate_sensitivity} are by definition independent from carbon
cycle--related effects, since both of them are defined in terms of a doubling
of the atmospheric \ac{CO2} concentration, not in terms of \ac{CO2} emissions.
Nevertheless, to assess policy-relevant metrics like the allowable fossil fuel
emissions to meet particular warming targets, for example, the $1.5
\unit{\degreeCelsius}$ of the Paris Agreement \autocite{UNFCCC2015}, it is
crucial to take the global carbon cycle into account. This
\namecref{sec:02:carbon_cycle} introduces the scientific background of the
global carbon cycle and its current anthropogenic perturbations.

@@

Thus, more sophisticated techniques are necessary to evaluate multi-model
climate projections. This \namecref{sec:02:reducing_uncertainties} introduces
three state-of-the-art methods to assess multi-model projections and reduce
associated uncertainties with observations. These techniques form the baseline
of the work presented in the following \namecrefs{ch:05:paper_ecs}.

@@

As indicated at the beginning of this \namecref{sec:02:reducing_uncertainties},
one main issue of multi-model ensembles is that not all participating climate
models are equally plausible. Usually, this is quantified with some kind of
measure of the models' \emph{performance}, \ie{} their agreement with
observations of the real climate system. However, this model performance can
only be evaluated against observations of the past and present-day climate,
which does not necessarily provide insights into the quality of model
projections of the future climate.

@@

Further techniques to reduce uncertainties in climate model projections with
observations are model weighting schemes. Their basic idea is to abandon model
democracy by replacing the arithmetic mean used to calculate the \acp{MMM} by a
weighted mean of the form
\begin{equation}
  y = \sum_{m=1}^{M} w_m y_m
  \label{eq:02:weighted_mean}
\end{equation}
with normalized weights $w_m$. Similar to the notation introduced in the
previous \namecref{subsec:02:emergent_constraints}, $y$ is a target variable
(\eg{} a projection of the future climate) and $m$ indexes the $M$ different
climate models. To address two major issues of model democracy (different
climate models are not equally plausible and not independent),
\textcite{Knutti2017a} propose a weighting scheme based on climate model
performance and interdependence with the following weights:
\begin{equation}
  w_m \propto \frac{\exp \left( -\frac{D_m^2}{\sigma_D^2} \right)}{1 + \sum_{n
    \neq m}^{M} \exp \left( -\frac{S_{mn}^2}{\sigma_S^2} \right)}.
  \label{eq:02:knutti_weights}
\end{equation}
The metric $D_m$ describes the distance between climate model $m$ and
observations (= model performance) and the metric $S_{mn}$ describes the
distance between climate model $m$ and $n$ (= model interdependence).
$\sigma_D$ and $\sigma_S$ are constants that determine the individual strength
of the performance and interdependence weighting, respectively.

@@

An alternative climate model weighting scheme is the \emph{\ac{MDER}} approach
\autocite{Karpechko2013}. Similar to all methods presented in
\cref{sec:02:reducing_uncertainties}, it can be used to reduce uncertainties in
climate model projections with observations. The basis of \ac{MDER} is a set of
$K$ predictor diagnostics $\left\{ x^{(1)}, x^{(2)}, \ldots, x^{(K)} \right\}$
which are relevant for the projection of the target variable $y$. The reasoning
for this choice of diagnostics is similar to the one presented in the previous
\namecref{subsec:02:model_weighting}: Weighting schemes should address the
question \enquote{which climate model is adequate for predicting the target
  variable?} and not \enquote{which climate model is the best?}.

>>>>

This \lcnamecref{ch:02:scientific_background} introduces the scientific
background of this thesis. First, basic concepts of climate model simulations
and associated uncertainties are introduced (\cref{sec:02:esms}). Next,
important metrics describing climate sensitivity
(\cref{sec:02:climate_sensitivity}) and fundamental biogeochemical processes of
the global carbon cycle (\cref{sec:02:esms}) are presented. Finally,
state-of-the-art techniques that can be used to reduce uncertainties in
projections of the future climate with observations are shown
(\cref{sec:02:reducing_uncertainties}). These methods form the basis for new
techniques developed in this thesis.

@@

In this thesis, climate model data from the two most recent \ac{CMIP}
generations is used, \acs{CMIP}5 and \acs{CMIP}6. More detailed information
about the specific variables and experiments analyzed is given in the
corresponding \lcnamecrefs{ch:04:papers_ecs_tcr_assessment}.

@@

Currently, only about $50 \unit{\%}$ of the emitted \ac{CO2} by humankind
remains in the atmosphere \autocite{Friedlingstein2020}. The residual part is
absorbed by the two other main carbon sinks of the planet: the terrestrial
biosphere and the ocean. These exchange fluxes of carbon are part of the global
carbon cycle of planet Earth. Since only atmospheric \ac{CO2} can act as
\ac{GHG}, this removal of \ac{CO2} from the atmosphere substantially slows down
global warming. Moreover, since the emission of carbon-based \acp{GHG}
introduces additional carbon into the Earth system, the carbon exchange fluxes
might change under global warming, which in turn directly influences the
magnitude of climate sensitivity. However, the two idealized climate
sensitivity metrics \ac{ECS} and \ac{TCR} introduced in the previous
\lcnamecref{sec:02:climate_sensitivity} are by definition independent from
carbon cycle--related effects, since both of them are defined in terms of a
doubling of the atmospheric \ac{CO2} concentration, not in terms of \ac{CO2}
emissions. Nevertheless, to assess policy-relevant metrics like the allowable
fossil fuel emissions to meet particular warming targets, for example, the $1.5
\unit{\degreeCelsius}$ of the Paris Agreement \autocite{UNFCCC2015}, it is
crucial to take the global carbon cycle into account. This
\lcnamecref{sec:02:carbon_cycle} introduces the scientific background of the
global carbon cycle and its current anthropogenic perturbations.

@@

Thus, more sophisticated techniques are necessary to evaluate multi-model
climate projections. This \lcnamecref{sec:02:reducing_uncertainties} introduces
three state-of-the-art methods to assess multi-model projections and reduce
associated uncertainties with observations. These techniques form the baseline
of the work presented in the following \lcnamecrefs{ch:05:paper_ecs}.

@@

As indicated at the beginning of this
\lcnamecref{sec:02:reducing_uncertainties}, one main issue of multi-model
ensembles is that not all participating climate models are equally plausible.
Usually, this is quantified with some kind of measure of the models'
\emph{performance}, \ie{} their agreement with observations of the real climate
system. However, this model performance can only be evaluated against
observations of the past and present-day climate, which does not necessarily
provide insights into the quality of model projections of the future climate.

@@

Further techniques to reduce uncertainties in climate model projections with
observations are model weighting schemes. Their basic idea is to abandon model
democracy by replacing the arithmetic mean used to calculate the \acp{MMM} by a
weighted mean of the form
\begin{equation}
  y = \sum_{m=1}^{M} w_m y_m
  \label{eq:02:weighted_mean}
\end{equation}
with normalized weights $w_m$. Similar to the notation introduced in the
previous \lcnamecref{subsec:02:emergent_constraints}, $y$ is a target variable
(\eg{} a projection of the future climate) and $m$ indexes the $M$ different
climate models. To address two major issues of model democracy (different
climate models are not equally plausible and not independent),
\textcite{Knutti2017a} propose a weighting scheme based on climate model
performance and interdependence with the following weights:
\begin{equation}
  w_m \propto \frac{\exp \left( -\frac{D_m^2}{\sigma_D^2} \right)}{1 + \sum_{n
    \neq m}^{M} \exp \left( -\frac{S_{mn}^2}{\sigma_S^2} \right)}.
  \label{eq:02:knutti_weights}
\end{equation}
The metric $D_m$ describes the distance between climate model $m$ and
observations (= model performance) and the metric $S_{mn}$ describes the
distance between climate model $m$ and $n$ (= model interdependence).
$\sigma_D$ and $\sigma_S$ are constants that determine the individual strength
of the performance and interdependence weighting, respectively.

@@

An alternative climate model weighting scheme is the \emph{\ac{MDER}} approach
\autocite{Karpechko2013}. Similar to all methods presented in
\cref{sec:02:reducing_uncertainties}, it can be used to reduce uncertainties in
climate model projections with observations. The basis of \ac{MDER} is a set of
$K$ predictor diagnostics $\left\{ x^{(1)}, x^{(2)}, \ldots, x^{(K)} \right\}$
which are relevant for the projection of the target variable $y$. The reasoning
for this choice of diagnostics is similar to the one presented in the previous
\lcnamecref{subsec:02:model_weighting}: Weighting schemes should address the
question \enquote{which climate model is adequate for predicting the target
  variable?} and not \enquote{which climate model is the best?}.

### 2021-04-09 ch03_esmvaltool/esmvaltool.tex
Used lowercase versions of namecref(s) at correct places

To address the \namecrefs{enum:01:question_1} of this thesis posed in
\cref{sec:01:key_science_questions}, a reliable and efficient tool to read,
process, and evaluate climate model output and observational data is necessary.
A valuable software package that fits these criteria is the \ac{ESMValTool}.
The \ac{ESMValTool} is an open-source community diagnostics and performance
metrics tool for the routine evaluation of \ac{ESM} output, which notably
facilitates the analysis of \ac{CMIP} models
(\url{https://www.esmvaltool.org}). For this reason, all analyses presented in
\cref{ch:04:papers_ecs_tcr_assessment,ch:05:paper_ecs,ch:06:paper_gpp} of this
thesis have been implemented into the \ac{ESMValTool}. Apart from that, further
substantial changes and additions to the code base of the \ac{ESMValTool} have
been contributed that improve the routine evaluation of climate models, which
is beneficial for the entire scientific community. This led to co-authorship in
the scientific documentation of the \ac{ESMValTool}, which is published in four
peer-reviewed studies \autocite{Eyring2020, Lauer2020, Righi2020, Weigel2020}.
After a brief overview of the \ac{ESMValTool}'s structure
(\cref{sec:03:esmvaltool}), this \namecref{ch:03:esmvaltool} presents all these
contributions that are not documented in other
\namecrefs{ch:04:papers_ecs_tcr_assessment} of this thesis and puts them into
context
(\cref{sec:03:contributions_to_esmvalcore,sec:03:contributions_to_esmvaltool}).

@@

In addition to the ported recipes, several new recipes and diagnostic scripts
have been contributed to the public \ac{ESMValTool} repository (see
\cref{tab:03:changes_to_esmvaltool} for a summary). Some of these include the
analyses performed in other \namecrefs{ch:04:papers_ecs_tcr_assessment} of this
thesis, namely \emph{recipe\_bock20jgr.yml}
(\cref{ch:04:papers_ecs_tcr_assessment}), \emph{recipe\_meehl20sciadv.yml}
(\cref{ch:04:papers_ecs_tcr_assessment}), \emph{recipe\_schlund20esd.yml}
(\cref{ch:05:paper_ecs}), and \emph{recipe\_schlund20jgr\_*.yml}
(\cref{ch:06:paper_gpp}). \emph{recipe\_ecs.yml} and \emph{recipe\_tcr.yml}
contain diagnostic scripts that allow the calculation of the climate metrics
\ac{ECS} and \ac{TCR} for arbitrary model output (see
\cref{sec:02:climate_sensitivity}), which is also used in
\cref{ch:04:papers_ecs_tcr_assessment,ch:05:paper_ecs}. Corresponding plots are
shown in \cref{fig:02:gregory_regression,fig:02:tcr}.
\emph{recipe\_cox18nature.yml} reproduces the analysis of \textcite{Cox2018},
who introduce an emergent constraint in \ac{ECS} based on a global temperature
variability metric $\Psi$. \Cref{fig:03:cox18:a} shows $\Psi$ over the past 85
years for \acs{CMIP}5 models and observations from HadCRUT4
\autocite{Morice2012}. Since high-sensitivity climate models exhibit higher
values of $\Psi$ than low-sensitivity models, an emergent relationship between
\ac{ECS} and $\Psi$ can be established (\cref{fig:03:cox18:b}). With
observations of $\Psi$, \ac{ECS} can eventually be constrained to $\left( 2.8
\pm 0.6 \right) \unit{K}$ ($66 \unit{\%}$ confidence range). The emergent
constraint by \textcite{Cox2018} is, together with 10 other emergent
constraints on \ac{ECS}, analyzed in great detail for the \acs{CMIP}5 and
\acs{CMIP}6 ensemble in \cref{ch:05:paper_ecs}, which also includes a
description of the calculation of $\Psi$ in \cref{subsec:05:cox}.

>>>>

To address the \lcnamecrefs{enum:01:question_1} of this thesis posed in
\cref{sec:01:key_science_questions}, a reliable and efficient tool to read,
process, and evaluate climate model output and observational data is necessary.
A valuable software package that fits these criteria is the \ac{ESMValTool}.
The \ac{ESMValTool} is an open-source community diagnostics and performance
metrics tool for the routine evaluation of \ac{ESM} output, which notably
facilitates the analysis of \ac{CMIP} models
(\url{https://www.esmvaltool.org}). For this reason, all analyses presented in
\cref{ch:04:papers_ecs_tcr_assessment,ch:05:paper_ecs,ch:06:paper_gpp} of this
thesis have been implemented into the \ac{ESMValTool}. Apart from that, further
substantial changes and additions to the code base of the \ac{ESMValTool} have
been contributed that improve the routine evaluation of climate models, which
is beneficial for the entire scientific community. This led to co-authorship in
the scientific documentation of the \ac{ESMValTool}, which is published in four
peer-reviewed studies \autocite{Eyring2020, Lauer2020, Righi2020, Weigel2020}.
After a brief overview of the \ac{ESMValTool}'s structure
(\cref{sec:03:esmvaltool}), this \lcnamecref{ch:03:esmvaltool} presents all
these contributions that are not documented in other
\lcnamecrefs{ch:04:papers_ecs_tcr_assessment} of this thesis and puts them into
context
(\cref{sec:03:contributions_to_esmvalcore,sec:03:contributions_to_esmvaltool}).

@@

In addition to the ported recipes, several new recipes and diagnostic scripts
have been contributed to the public \ac{ESMValTool} repository (see
\cref{tab:03:changes_to_esmvaltool} for a summary). Some of these include the
analyses performed in other \lcnamecrefs{ch:04:papers_ecs_tcr_assessment} of
this thesis, namely \emph{recipe\_bock20jgr.yml}
(\cref{ch:04:papers_ecs_tcr_assessment}), \emph{recipe\_meehl20sciadv.yml}
(\cref{ch:04:papers_ecs_tcr_assessment}), \emph{recipe\_schlund20esd.yml}
(\cref{ch:05:paper_ecs}), and \emph{recipe\_schlund20jgr\_*.yml}
(\cref{ch:06:paper_gpp}). \emph{recipe\_ecs.yml} and \emph{recipe\_tcr.yml}
contain diagnostic scripts that allow the calculation of the climate metrics
\ac{ECS} and \ac{TCR} for arbitrary model output (see
\cref{sec:02:climate_sensitivity}), which is also used in
\cref{ch:04:papers_ecs_tcr_assessment,ch:05:paper_ecs}. Corresponding plots are
shown in \cref{fig:02:gregory_regression,fig:02:tcr}.
\emph{recipe\_cox18nature.yml} reproduces the analysis of \textcite{Cox2018},
who introduce an emergent constraint in \ac{ECS} based on a global temperature
variability metric $\Psi$. \Cref{fig:03:cox18:a} shows $\Psi$ over the past 85
years for \acs{CMIP}5 models and observations from HadCRUT4
\autocite{Morice2012}. Since high-sensitivity climate models exhibit higher
values of $\Psi$ than low-sensitivity models, an emergent relationship between
\ac{ECS} and $\Psi$ can be established (\cref{fig:03:cox18:b}). With
observations of $\Psi$, \ac{ECS} can eventually be constrained to $\left( 2.8
\pm 0.6 \right) \unit{K}$ ($66 \unit{\%}$ confidence range). The emergent
constraint by \textcite{Cox2018} is, together with 10 other emergent
constraints on \ac{ECS}, analyzed in great detail for the \acs{CMIP}5 and
\acs{CMIP}6 ensemble in \cref{ch:05:paper_ecs}, which also includes a
description of the calculation of $\Psi$ in \cref{subsec:05:cox}.

### 2021-04-09 ch04_papers_ecs_tcr_assessment/papers_ecs_tcr_assessment.tex
Used lowercase versions of namecref(s) at correct places

To reduce uncertainties in multi-model climate projections, a first important
step is the assessment of the desired target variables in the corresponding
climate model ensemble. This is particularly relevant when a new generation of
\acp{ESM} is published that includes considerable modifications compared to the
respective predecessor model versions. Currently, we are in a situation like
this with the \acs{CMIP}6 ensemble, for which new data is still released every
day (as of \TheMonth{}). In light of the upcoming international climate
assessment of the \ac{IPCC}'s \acs{AR}6, the evaluation of policy-relevant
metrics like \ac{ECS} or \ac{TCR} for these new \acs{CMIP}6 models is crucial
since they provide vital information about the future climate of the planet
Earth. As the \ac{ESMValTool} is an excellent tool that allows for a quick and
robust evaluation of new \ac{CMIP} data as soon as it gets published on the
\ac{ESGF} servers (see \cref{sec:03:esmvaltool}), it is used here to assess and
analyze \ac{ECS} and \ac{TCR} of the \acs{CMIP}6 models and compare it to
corresponding results of predecessor model generations and international
climate assessments. This work is already published in two scientific papers:
\textcite{Bock2020} and \textcite{Meehl2020}. For \textcite{Bock2020}, the
author of this thesis contributed their figure 8 (a bar chart showing the
\ac{ECS} for several \acs{CMIP}3, \acs{CMIP}5, and \acs{CMIP}6 models), their
figure 10 (map plots showing cloud-related feedback parameters for the
\acs{CMIP}5 and \acs{CMIP}6 \acp{MMM}; see
\cref{fig:04:cloud_feedback_parameters}), code to create these figures with the
\ac{ESMValTool}, and text for the manuscript (in particular a section about the
evaluation of \ac{ECS}). For \textcite{Meehl2020}, the author of this thesis
contributed all figures and tables, code to produce these with the
\ac{ESMValTool}, and text for the manuscript. This
\namecref{ch:04:papers_ecs_tcr_assessment} first presents the evaluation of
climate sensitivity for the \acs{CMIP}5 and \acs{CMIP}6 ensembles
(\cref{sec:04:evaluation_ecs_and_tcr}) and puts the corresponding \ac{ECS} and
\ac{TCR} values into historical context (\cref{sec:04:historical_context}).
Finally, possible reasons for the apparent increase of climate sensitivity in
the \acs{CMIP}6 models are discussed in detail
(\cref{sec:04:possible_reasons_high_ecs_cmip6}).

@@

To illustrate the results of the previous
\namecref{sec:04:evaluation_ecs_and_tcr} and put them into historical context,
\cref{fig:04:historical_ecs_tcr_ranges} shows the assessed \ac{ECS} and
\ac{TCR} ranges over the years from the Charney report \autocite{Charney1979}
and the different \acp{AR} of the \ac{IPCC} in combination with the
corresponding modeled ranges from the different \ac{CMIP} generations. Since
the Charney report in 1979, the assessed range of \ac{ECS} of
$\rangeunit{1.5}{4.5}{K}$ has almost remained unchanged during the last 40
years \autocite{Charney1979, Mitchell1990, Kattenberg1996, Albritton2001,
  Stocker2013}, with the exception of \acs{AR}4, where the lower bound was
temporarily increased to $2.0 \unit{K}$ \autocite{Solomon2007}. For \acs{AR}2,
\acs{AR}4, and \acs{AR}5, the corresponding climate model generations from
\acs{CMIP}1, \acs{CMIP}3, and \acs{CMIP}5 more or less agree with this assessed
range \autocite{Kattenberg1996, Randall2007, Flato2013}. In contrast to that,
some of the early climate models used in \acs{AR}1 and some \acs{CMIP}2 models
used in \acs{AR}3 exhibit values of \acs{ECS} well above $4.5 \unit{K}$,
resulting in an upper model range of about $0.6 \unit{K}$ higher than the
assessed range \autocite{Mitchell1990, Cubasch2001}. However, these deviations
from the assessed range are small compared to the ones present in the
\acs{CMIP}6 ensemble, which have become evident over the course of 2020. As
shown in \cref{fig:04:historical_ecs_tcr_ranges}, both the lower and upper
bound are more extreme than in any previous climate model generation, resulting
in a \acs{CMIP}6 \ac{ECS} model range of $\rangeunit{1.8}{5.6}{K}$ (as of
\TheMonth{}). This increase in range is particularly relevant for the upper
bound, which is about $1.1 \unit{K}$ higher than the assessed upper bound of
$4.5 \unit{K}$. Moreover, it is not only a small fraction of models that exceed
this upper bound: out of the 42 analyzed \acs{CMIP}6 models, a third (14
models) have \iac{ECS} above $4.5 \unit{K}$. In addition to the analysis
presented here, several modeling centers independently confirmed high \ac{ECS}
values in their models \autocite{Andrews2019, BodasSalcedo2019, Gettelman2019,
  Wyser2020}. A list of possible reasons for the increased climate sensitivity
in the \acs{CMIP}6 ensemble including aerosol-cloud interactions and changes in
the shortwave cloud feedback over the Southern Ocean is given in
\cref{sec:04:possible_reasons_high_ecs_cmip6}.

@@

Because of the massive political and societal relevance of \ac{ECS}, its
apparent increase in the \acs{CMIP}6 ensemble is currently one of the most
important questions for the entire climate modeling community. For this reason,
this \namecref{sec:04:possible_reasons_high_ecs_cmip6} discusses possible
reasons for the increased climate sensitivity in \acs{CMIP}6.

>>>>

To reduce uncertainties in multi-model climate projections, a first important
step is the assessment of the desired target variables in the corresponding
climate model ensemble. This is particularly relevant when a new generation of
\acp{ESM} is published that includes considerable modifications compared to the
respective predecessor model versions. Currently, we are in a situation like
this with the \acs{CMIP}6 ensemble, for which new data is still released every
day (as of \TheMonth{}). In light of the upcoming international climate
assessment of the \ac{IPCC}'s \acs{AR}6, the evaluation of policy-relevant
metrics like \ac{ECS} or \ac{TCR} for these new \acs{CMIP}6 models is crucial
since they provide vital information about the future climate of the planet
Earth. As the \ac{ESMValTool} is an excellent tool that allows for a quick and
robust evaluation of new \ac{CMIP} data as soon as it gets published on the
\ac{ESGF} servers (see \cref{sec:03:esmvaltool}), it is used here to assess and
analyze \ac{ECS} and \ac{TCR} of the \acs{CMIP}6 models and compare it to
corresponding results of predecessor model generations and international
climate assessments. This work is already published in two scientific papers:
\textcite{Bock2020} and \textcite{Meehl2020}. For \textcite{Bock2020}, the
author of this thesis contributed their figure 8 (a bar chart showing the
\ac{ECS} for several \acs{CMIP}3, \acs{CMIP}5, and \acs{CMIP}6 models), their
figure 10 (map plots showing cloud-related feedback parameters for the
\acs{CMIP}5 and \acs{CMIP}6 \acp{MMM}; see
\cref{fig:04:cloud_feedback_parameters}), code to create these figures with the
\ac{ESMValTool}, and text for the manuscript (in particular a section about the
evaluation of \ac{ECS}). For \textcite{Meehl2020}, the author of this thesis
contributed all figures and tables, code to produce these with the
\ac{ESMValTool}, and text for the manuscript. This
\lcnamecref{ch:04:papers_ecs_tcr_assessment} first presents the evaluation of
climate sensitivity for the \acs{CMIP}5 and \acs{CMIP}6 ensembles
(\cref{sec:04:evaluation_ecs_and_tcr}) and puts the corresponding \ac{ECS} and
\ac{TCR} values into historical context (\cref{sec:04:historical_context}).
Finally, possible reasons for the apparent increase of climate sensitivity in
the \acs{CMIP}6 models are discussed in detail
(\cref{sec:04:possible_reasons_high_ecs_cmip6}).

@@

To illustrate the results of the previous
\lcnamecref{sec:04:evaluation_ecs_and_tcr} and put them into historical
context, \cref{fig:04:historical_ecs_tcr_ranges} shows the assessed \ac{ECS}
and \ac{TCR} ranges over the years from the Charney report
\autocite{Charney1979} and the different \acp{AR} of the \ac{IPCC} in
combination with the corresponding modeled ranges from the different \ac{CMIP}
generations. Since the Charney report in 1979, the assessed range of \ac{ECS}
of $\rangeunit{1.5}{4.5}{K}$ has almost remained unchanged during the last 40
years \autocite{Charney1979, Mitchell1990, Kattenberg1996, Albritton2001,
  Stocker2013}, with the exception of \acs{AR}4, where the lower bound was
temporarily increased to $2.0 \unit{K}$ \autocite{Solomon2007}. For \acs{AR}2,
\acs{AR}4, and \acs{AR}5, the corresponding climate model generations from
\acs{CMIP}1, \acs{CMIP}3, and \acs{CMIP}5 more or less agree with this assessed
range \autocite{Kattenberg1996, Randall2007, Flato2013}. In contrast to that,
some of the early climate models used in \acs{AR}1 and some \acs{CMIP}2 models
used in \acs{AR}3 exhibit values of \acs{ECS} well above $4.5 \unit{K}$,
resulting in an upper model range of about $0.6 \unit{K}$ higher than the
assessed range \autocite{Mitchell1990, Cubasch2001}. However, these deviations
from the assessed range are small compared to the ones present in the
\acs{CMIP}6 ensemble, which have become evident over the course of 2020. As
shown in \cref{fig:04:historical_ecs_tcr_ranges}, both the lower and upper
bound are more extreme than in any previous climate model generation, resulting
in a \acs{CMIP}6 \ac{ECS} model range of $\rangeunit{1.8}{5.6}{K}$ (as of
\TheMonth{}). This increase in range is particularly relevant for the upper
bound, which is about $1.1 \unit{K}$ higher than the assessed upper bound of
$4.5 \unit{K}$. Moreover, it is not only a small fraction of models that exceed
this upper bound: out of the 42 analyzed \acs{CMIP}6 models, a third (14
models) have \iac{ECS} above $4.5 \unit{K}$. In addition to the analysis
presented here, several modeling centers independently confirmed high \ac{ECS}
values in their models \autocite{Andrews2019, BodasSalcedo2019, Gettelman2019,
  Wyser2020}. A list of possible reasons for the increased climate sensitivity in
the \acs{CMIP}6 ensemble including aerosol-cloud interactions and changes in
the shortwave cloud feedback over the Southern Ocean is given in
\cref{sec:04:possible_reasons_high_ecs_cmip6}.

@@

Because of the massive political and societal relevance of \ac{ECS}, its
apparent increase in the \acs{CMIP}6 ensemble is currently one of the most
important questions for the entire climate modeling community. For this reason,
this \lcnamecref{sec:04:possible_reasons_high_ecs_cmip6} discusses possible
reasons for the increased climate sensitivity in \acs{CMIP}6.
